Predicting Mental Workload: A Multimodal Intermediate Fusion Deep Learning Approach

### 2021-04-26 0.manuscript/manuscript.tex
Repo restructure + updates readme files

%Prerequisite statements
\DeclareCaptionLabelSeparator*{spaced}{\\[2ex]} %Declare the caption and title seperator spacing
\captionsetup[table]{,textfont=it,format=plain,justification=justified,
  singlelinecheck=false,labelsep=spaced,skip=2ex} %Setting up caption for table titles
 \captionsetup[figure]{,textfont=it,format=plain,justification=justified,
  singlelinecheck=false,labelsep=spaced,skip=2ex} %Setting up caption for figure titles
\captionsetup{labelfont={bf}} %Bold label font for tables
\onehalfspacing %Document line spacing
\pagestyle{myheadings} %Heading style
\graphicspath{ {./figures/} } %Defining image path

%Frontpage specifications 
\title{Predicting  Mental Workload: An Intermediate Fusion Deep Learning Approach}

\author[1]{Bart-Jan Boverhof \thanks{b.boverhof@students.uu.nl}}
\author[2]{Bernard P. Veldkamp\thanks{b.p.veldkamp@utwente.nl}}
\affil[1]{\normalsize Faculty of Behavioral, Management and Social Sciences, University of Twente}
\affil[2]{\normalsize Faculty of Social and Behavioural Sciences, Utrecht University}

@@

\begin{document}

\maketitle

@@

\hspace*{3.6mm}\textit{Keywords:} lorem, ipsum, dolor, sit amet, lectus % Keywords

@@

\begin{equation}
\label{eq:corr}
r_{X,Y} = \frac{cov(X,Y)}{\sigma_{X} \sigma_{Y}}
\end{equation}
where $x$ refers to the predicted values for the test windows,  $y$ to the true labels of test windows and $\sigma_{X}\, \& \,\sigma_{Y}$ refer to the standard deviation of $x$ and $y$ respectively.  Conceptually, correlation indicates the coherence between predictions and labels, hereby posing an alternative model performance metric.

@@

\newpage
\section{Conclusion and Discussion}
We found that the unimodal EEG architecture consistently scored best on all performance metrics,  with a rather sizable difference when considering correlations between predictions and labels. Rather unexpected is the inadequate performance of the multimodal architecture. One is enticed to expect that the architecture combining information from multiple modalities yields a multifaceted, hence richer, assessment of workload, and as a consequence attains more adequate performance. Indeed this is also to be distinguished from previous research, lending credence to the unexpected character of our results \cite{dolmans2020perceived} \cite{han2020classification} \cite{rastgoo2019automatic}. Furthermore, score on performance metrics for the GSR architecture are consistently lower as compared with the EEG architecture, but consistently higher as compared with the multimodal architecture. In terms of absolute errors, these differences were not found to be statistically significant, however.  The unimodal PPG architecture was found to perform worst across the board.

Although some differences in performance are distinguishable, one must recognize that these differences are of rather modest size. Some of the differences in absolute error, although being statistically significant, still differ virtually negligibly in terms of their absolute difference, as indicated by both MAE and RMSE. The statistical significance is particularly attributable to the vast amount of degrees of freedom with which was tested. Nevertheless, the fact that the multimodal architecture didn't outperform a single of the unimodel architectures deserves some further investigation.

One imaginable train of thought is that the multimodal architecture should have outperformed at least some of the unimodal architectures, but this was stiffed by the head-network architecture. Despite some different head-networks were cross-compared, this was not done in a systematical manner, i.e. such as would be possible with HPO. We finally opted for a relatively shallow head-network, consisting of several fully connected dense layers. The reader might recall from Section \ref{section:multimodal} that HPO and training for the current multimodal architecture already demanded hefty computational resources, and thus considerable time. Utilizing a deeper head-network would increase the required computational resources and hence running time exceedingly more.  Additionally, in spite of some previous research utilizing a relatively deep head-network, a rather shallow network was equally so proposed oftentimes for multimodal workload classification problems \cite{yin2017recognition}, \cite{han2020classification} \cite{rastgoo2019automatic}. Moreover, one could propose that that the performance of the multimodal architecture was stiffed by the poor performing unimodal PPG component. In order to scrutinize this, we optimized hyperparamaters and trained a multimodal architecture in exactly the same manner, except for omitting the PPG component altogether. This network did not perform better on any metric, and results were not reported as a consequence.

A contrary train of thought is that our case simply didn't benefit from a multimodal approach.  A classical theorem in the fields of statistics and machine-learning alike is the, sometimes precarious, balance between model complexity and and simplicity. A model that is overly complex is in increased peril of overfitting, i.e. performing well on training data, but failing to generalize to the test data.  A more parsimonious model is generally at less jeopardy to fall into this pitfall, however an overly simple model simply doesn't describe the data very well. It is not implausible to consider that this mechanism could, at least, partially rationalize why our relatively complex multimodal architecture performs worse as compared with at least one of the simpler unimodal architecture. The fact that we trained an individual network for each participant separately, inherently implying that training is not done on an extraordinary sizable amount of data, may lend credence to this explanation: training on insufficient data equally so increases the likelihood of overfitting.

One is therefore inclined to reflect on the following: Even if a multimodal approach could result in a moderate improvement over unimodal approaches, would this justify the sheer addition in complexity,  the considerable increase in required computational resources and the necessity to possess a multitude of different physiological sensors? The answers to this question is naturally heavily dependent upon context, i.e. how precise need the network be for the task at hand and how much funding and time is available? It is not

The two previously delineated opposite lines of argumentation can be distilled into recommendations for future research endeavors. Firstly, insights into the most appropriate network architecture of, in particular, the multimodal head-network is would be beneficial. Despite the range of different proposals to architectures with the purpose of workload classification by amongst others \citeA{yin2017recognition}, \citeA{han2020classification} \citeA{rastgoo2019automatic} and \citeA{dolmans2020perceived}, some rough outlines into what is prone to work under different circumstances would be highly beneficial.

\begin{itemize}
  \item  More HPO could have been done, to get even better performing models but we had limited computatioal power.   
  \item This approach costs a lot of computational power to both train and do HPO. Especially since person-specific models are fitted. This upholds mostly for the multimodal network.
  \item Due to the sheer size of the trained model and the fact that models are trained person specific, storage sizes of the models are huge. This could be unpractical.  This upholds mostly for the multimodal network.
  \item Possibly different window sizes per modality could lead to better results.  * Insert some of the sources here that suggest different window sizes per modality* . This was beyond the scope of this project.
   \item How generizable is this to other situations for the same person (since things such as PPG are context dependent (like for example regarding how one ate).
   \item Long runtimes,  even when testing. This upholds only for the multi network. The reason is likely the sheer size of these networks, which for some persons are blown up due to device lagging. 
   \item hardware failure?
   \item The fact that we use TLX as labels.
\end{itemize}

>>>>

%Prerequisite statements
\DeclareCaptionLabelSeparator*{spaced}{\\[2ex]} %Declare the caption and title seperator spacing
\captionsetup[table]{,textfont=it,format=plain,justification=justified,
  singlelinecheck=false,labelsep=spaced,skip=2ex} %Setting up caption for table titles
 \captionsetup[figure]{,textfont=it,format=plain,justification=justified,
  singlelinecheck=false,labelsep=spaced,skip=2ex} %Setting up caption for figure titles
\captionsetup{labelfont={bf}} %Bold label font for tables
\onehalfspacing %Document line spacing
\pagestyle{myheadings} %Heading style
\graphicspath{ {./figures/} } %Defining image path
\pagestyle{myheadings}
\markright{A Deep Learning Approach to Mental Workload Prediction}

%Frontpage specifications 
\title{Predicting  Mental Workload: A Multimodal Intermediate Fusion Deep Learning Approach}

\author[1]{Bart-Jan Boverhof \thanks{b.boverhof@students.uu.nl}}
\author[2]{Bernard P. Veldkamp}
\affil[1]{\normalsize Faculty of Social and Behavioural Sciences, Utrecht University}
\affil[2]{\normalsize Faculty of Behavioral, Management and Social Sciences, University of Twente}

@@

\begin{document}
\maketitle
\thispagestyle{empty}

@@

\vspace{1cm}
\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
\textit{Keywords:} deep-learning, multimodal, mental workload, intermediate fusion, hyperparameter optimization (HPO), brain-computer Interface (\textit{BCI}),  electroencephalography (\textit{EEG}), galvanic skin response (\textit{GSR}), photoplethysmography (\textit{PPG})
\end{minipage}

@@

\begin{equation}
\label{eq:corr}
r_{X,Y} = \frac{cov(X,Y)}{\sigma_{X} \sigma_{Y}}
\end{equation}
where $x$ refers to the predicted values for the test windows,  $y$ to the true labels of test windows, $cov(X,Y)$ to their covariance, and $\sigma_{X}\, \& \,\sigma_{Y}$ refer to the standard deviation of $x$ and $y$ respectively.  Conceptually, correlation indicates the coherence between predictions and labels, hereby posing an alternative model performance metric.

@@

\newpage
\section{Discussion \& Conclusion}
\subsection{Discussion Results}
We found the unimodal EEG architecture to consistently surpass the other architectures in terms of their scores on the performance metrics,  with a rather sizable difference in the correlations between predictions and labels.  Although no statistically significant difference in absolute error was found between the unimodal EEG and GSR architectures, we did find evidence in favor of the unimodal EEG architecture significantly outperforming the multimodel architecture. Furthermore, the performance of GSR architecture was detected to be consistently worse as compared with the EEG architecture, but consistently better as compared with the multimodal architecture, however in terms of absolute error these differences were not found to be statistically significant. The unimodal PPG architecture was found to perform worst across the board, however in terms of absolute error this difference was not found to be statistically significant when evaluated against the multimodal architecture.

Perhaps the most noteworthy result is that the unimodel EEG architecture outperformed the multimodal architecture. One is gravitated to believe that, since the architecture combining information from multiple modalities yields a multifaceted, hence richer, assessment of workload, performance would be more formidable. Indeed, this can be distinguished from previous research, lending credence to the unexpected character of our results \cite{dolmans2020perceived} \cite{han2020classification} \cite{rastgoo2019automatic} \cite{yin2017recognition}.  We will aim to rationalize these results in the subsequent sections.

One imaginable train of thought constitutes that the multimodal architecture should have outperformed (at least some of) the unimodal architectures, but this was stiffed by the relatively simple head-network architecture. Despite that some different head-network architectures were considered, no systematical comparison was conducted to select the best performing architecture. We opted for a relatively shallow head-network architecture.  However, the most important reason for opting for a relatively shallow head-network architecture was that a rather shallow network was found to be successful in many previous multimodal workload classification problems \cite{yin2017recognition}, \cite{han2020classification} \cite{rastgoo2019automatic}. Additionally, utilizing a deeper head-network would have substantially increased the computational expenses (which were already of quite considerable size) even further.

Further investigation is therefore pursued along a contrary line of reasoning, in which we propose that our case simply didn't benefit from a multimodal approach.  A classical theorem in the fields of statistics and machine-learning alike, is the oftentimes precarious balance between model complexity and simplicity. A model that is overly complex is in increased peril of overfitting, i.e. performing well on training data, but failing to generalize this adequate prediction towards the testing data.  With a simpler model, one is generally at less jeopardy to step into this pitfall, however an overly simplistic model is simply inaccurate \cite{lever2016points}. It is not implausible that this mechanism could, at least partially, rationalize that our rather complex multimodal architecture performed worse as compared with the considerably more parsimonious unimodal architecture. Lending more credence to this justification is the fact that we trained an individual network for each participant, inherently implying that training is done on a relatively small amount of data: this equally so increases the risk of overfitting.

Another noteworthy result was the delineated relationship between a window's absolute error and its respective label score (as was visualized by Figure \ref{fig:mae_scatter}), which indicated that all architectures tended to underpredict high labeled windows.  A credible explanation for this anomaly is the relatively minor presence of high labeled windows in our data (as was visualized by Figure \ref{fig:labels}). Naturally, when a relatively small amount of high labeled windows, i.e. high workload windows, are present in the training data, the DNN doesn't adequately learn to predict such windows.

Concluding the discussion section, one should note that, despite a difference in performance across architectures was observed,  these differences were of rather modest size. Some of the differences in absolute error, although being statistically significant, are still very small in terms of their absolute difference, as indicated by how close both MAE and RMSE scores were observed to be across architectures. The statistical significance of some of the conducted t-tests can particularly be attributed to the vast amount of degrees of freedom with which was tested. We therefore suggest to reflect on the following: in opposite to our finding, suppose that if a multimodal approach could result in a moderate improvement over an unimodal approach. Would this justify the sheer addition in complexity,  the considerable increment in computational expenditure and the necessity to invest in a multitude of different physiological sensors? In other words, isn't the already particularly adequately performing unimodal EEG approach sufficient for most task at hand? The answers to this question is naturally heavily context dependent, and should be considered in a case-specific manner.

\subsection{Limitations \& Future Research Recommendations}
More research is necessary to be able to answer the question concluding the previous section. Can the fact that we didn't find the multimodel architecture to perform superiourly,  in opposite to previous research, be attributed to the too shallow (head-network) DNN architecure? If yes,  what would be an appropriate architecture? Or is the current multimodal approach to complex for the data at hand, resulting in overfitting? An effective technique with which to answers these questions is HPO.  With HPO, different network architectures can be contrasted, with the objective of systematically finding the best performing architecture. In the current study we did optimize several hyperparamaters, however the architecture in itself was set in stone.  The reason for doing so were both restrictions in available time and computational power. A recommendation for future research is therefore to further investigate the most suitable DNN architecture for multimodal approaches.

As was delineated before, another noteworthy result was the underprediction  of high labeled windows, most likely due to their relative absence in training data. This absence naturally stems from the fact that respondents refrained from providing extremely high answers towards the items of the NASA-TLX questionnaire, which were utilized for labeling. Alternative labeling should therefore be considered in future research endeavors. Ideally, one would want to contrast various labeling strategies with one another, and derive those most suitable for the task at hand. Again, due to restrictions in available time and computational power, this lay outside the scope of the current research. Potentially interesting alternatives approaches to labeling could for example include a standardized version of the NASA-TLX, or an objective difficulty score of the task at hand as proposed by \citeA{dolmans2020perceived}, or lastly, an extensive combination of personal information such as big-five personality scale, sex, age, etc. as proposed by \citeA{sun2019hybrid}.

Finally, as was elaborated on before, the ultimate goal of this line of research is obtain trained networks with which workload assessment can be done in simulational scenario's, with the objective of training its participants. In order to acknowledge the different character of physiological data across participants, a separate network for each individual was trained.  However, apart from between person differences in physiological data, differences between occasions could equally so prevail. An example of such a circumstantial difference could entail that a person has a higher baseline stress level in one occasion over the other, which could imply that a personally trained DNN is not generalizable to another occasion.  Other of such imaginable circumstantial factors that could interfere with physiological signals includes food consumption before the simulation and room temperature.  Further research is required in order to assess the generelizability of a single personally trained DNN across different occasions, and is suggested as a consequence.

\subsection{Conclusion}
We embarked on a multimodal deep learning approach to workload prediction, by combining three physiological sensors, being EEG, GSR and PPG, into a single deep neural network. An intermediate fusion approach was adopted towards the construction of the multimodal architecture. The performance of this approach was contrasted with three unimodal approaches, each utilizing one of the previously delineated modalities. It was found that all architectures performed rather adequately. A difference in performance was observable however: most notably it was found that the unimodal EEG architecture outperformed the multimodal architecture. Several explanations were proposed for this anomaly, which were distilled into various recommendations for future research. Future work will ideally investigate upon the most appropriate network architecture for multimodal approaches,  the feasibility of different labeling strategies and the cross-occasional generelization of individually trained networks.
